---
title: "Hands-on Exercise 4b - Visual Statistical Analysis"
author: "Goh Si Hui"
date: 2024/02/01
date-format: long
date-modified: "last-modified"
format: html 
execute: 
  echo: true
  eval: true
  warning: false
editor: visual 
---

# About this Exercise

In this hands-on exercise, we will learn how to:

-   create visual graphics with rich statistical information using ggstatsplot package

-   visualise model diagnostics using performance package

-   visualise model parameters using parameters package.

# Getting Started

Before we start, let us ensure that the required R packages have been installed and import the relevant data for this hands-on exercise.

## Installing and Loading the Packages

For this exercise, other than tidyverse, we will use the following packages:

-   ggstatplot: an extension of ggplot2 package for creating graphics with details from statistical tests included in the information-rich plots themselves.

-   performance: to provide utilities for computing indices of model quality and goodness of fit.

-   parameters: to provide utilities for processing the parameters of various statistical models

-   tidyverse: a family of R packages for data science processing

-   readxl: to import excel files into R

The code chunk below uses `p_load()` of **pacman** package to check if the abovementioned packages are installed in the computer. If they are, they will be launched in R. Otherwise, **pacman** will install the relevant packages before launching them.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
pacman::p_load(ggstatsplot, performance, parameters, tidyverse, readxl, see)

```

## Importing the Data

For this exercise, we will be using the Exam_data.csv provided by the course instructor and we have used it in Hands-on Exercises 1 and 2. It consists of year end examination grades of a cohort of primary 3 students from a local school. It is in csv file format.

We use `read_csv()` function of **readr** to import the ***Exam_data.csv*** file into R and save it as a tibble data frame called `exam_data`. Then we will use `datatable()` of **DT** to have an overview of the imported data.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
exam_data <- read_csv("data/Exam_data.csv")
glimpse(exam_data)

```

In addition, we will be using the ***ToyotaCorolla.xls*** for Visualising Models and Parameters. We use `read_xls()` of readxl package to import the data worksheet of **ToyotaCorolla.xls** workbook into R.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
car_resale <- read_xls("data/ToyotaCorolla.xls", 
                       "data")
glimpse(car_resale)

```

# Visual Statistical Analysis

## One-sample test: gghistostats() method

We can use `gghistostats()` to build an visual of one-sample test on English Scores.

```{r}
set.seed(2024)

gghistostats(data = exam_data, x = ENGLISH, 
             type = "bayes",
             test.value = 60, 
             xlab= "English Scores")

```

::: callout-note
The default information presented are:

-   Statistical details

-   Bayes Factor

-   Sample Sizes

-   Distribution Summary
:::

### Unpacking the Bayes Factor

-   A Bayes factor is the ratio of the likelihood of one particular hypothesis to the likelihood of another. It can be interpreted as a measure of the strength of evidence in favor of one theory among two competing theories.

-   Bayes factor gives us a way to evaluate the data in favor of a null hypothesis, and to use external information to do so. It tells us what the weight of the evidence is in favor of a given hypothesis.

-   When we are comparing two hypotheses, H1 (the alternate hypothesis) and H0 (the null hypothesis), the Bayes Factor is often written as B10.

-   The Schwarz criterion is one of the easiest ways to calculate rough approximation of the Bayes Factor.

-   A Bayes Factor can be any positive number (i.e., 0 and ∞).

    -   1 indicates the data do not favor either theory more than the other;

    -   values greater than 1 indicate increasing evidence for one theory over the other (e.g., the alternative over a null hypothesis) ; and

    -   values less than 1 the converse (e.g., increasing evidence for the null over the alternative hypothesis).

-   Thus, Bayes factors allow three different types of conclusions:

    -   There is strong evidence for the alternative (B much greater than 1);

    -   there is strong evidence for the null (B close to 0); and

    -   the evidence is insensitive (B close to 1).

## Two-sample mean test: ggbetweenstats()

We can use `ggbetweenstats()` to build an visual of two sample means test of math scores by gender.

```{r}

ggbetweenstats(data = exam_data, x = GENDER, 
               y = MATHS, type = "np", 
               messages = FALSE)

```

::: callout-note
The default information presented are:

-   Statistical details

-   Bayes Factor

-   Sample Sizes

-   Distribution Summary
:::

## One Way ANOVA Test: ggbetweenstats() method

We can use `ggbetweenstats()` to build a visual for one-way ANOVA test on English Score by race.

```{r}
ggbetweenstats(
  data = exam_data, x = RACE, y = ENGLISH, 
  type = "p", mean.ci = TRUE, 
  pairwise_comparisons = TRUE, 
  pairwise.display = "s",
  p.adjust.method = "fdr", 
  messages = FALSE)


```

Try: explore the aov object!

## Significant Test of Correlation: ggscatterstats()

We can use `ggscatterstats()` to build a visual for significant test of correlation between Math Scores and English Scores.

```{r}
ggscatterstats(data = exam_data, 
               x = MATHS,
               y = ENGLISH, 
               marginal = FALSE)

```

## Significant Test of Association (Dependence): ggbarstats()

We can use `ggbarstats()` to build a visual for significant test of association.

For our data, we will first bin the Math scores into a 4-class variable using `cut()`

```{r}
exam1 <- exam_data %>%
  mutate(MATHS_bins = cut(MATHS, breaks = c(0, 60,75,85,100)))
```

Then we will build the visual using `ggbarstats()`.

```{r}
ggbarstats(exam1, x = MATHS_bins, 
           y = GENDER)

```

# Visualising Model Diagnostic and Model Parameters

## Multiple Regression Model using lm()

The following code chunk is used to calibrate a multiple linear regression model using `lm()` of Base Stats of R.

```{r}
model <- lm(Price ~ Age_08_04 + Mfg_Year + KM + 
              Weight + Guarantee_Period, data = car_resale)
model
```

## Model Diagnostic: Checks for Multicolinearity

We check for multicolinearity using `check_collinearity()` function of performance package.

```{r}
check_collinearity(model)
```

```{r}
check_c <- check_collinearity(model)
plot(check_c)
```

## Model Diagnostic: Check for Normality

We use `check_normality()` of performance package to check if the model follows the normality assumption.

```{r}
model1 <- lm(Price ~ Age_08_04 + KM + 
              Weight + Guarantee_Period, data = car_resale)

check_n <- check_normality(model1)

plot(check_n)
```

## Model Diagnostic: Check for Homogeneity of Variances

We check for homogeneity of Variances using `check_heterscedasticity` of performance package.

```{r}

check_h <- check_heteroscedasticity(model1)
plot(check_h)
```

## Model Diagnostic: Complete Check

We can also perform the complete by using `check_model()`.

```{r fig.width=12, fig.height=10}
check_model(model1)

```

## Visualising Regression Parameters

We can use two methods to visualise regression parameters: (i) see method and (ii) ggcoefstats() method.

::: panel-tabset
## see method

```{r}
plot(parameters(model1))

```

## ggcoefstats() method

```{r}
ggcoefstats(model1, 
            output = "plot")

```
:::

# References

-   Kam, T. S. (2023). *R for Visual Analytics* \[Web-book\]. <https://r4va.netlify.app/>.
