---
title: "Take-home Exercise 4: Prototyping Time Series Module for Visual Analytics Shiny Application" 
author: "Goh Si Hui"
date: 2024/02/25
execute: 
  echo: true
  eval: true
  freeze: true
  warning: false
  message: false
---

# About this Exercise
In this exercise, we will be preparing the time-series forecasting module of the proposed Shiny Application and complete the following task:

-   Evaluate and determine the necessary R packages needed for my group project's Shiny application;

-   Prepare and test the specific R codes can be run and returned the correct output as expected; 

-   Determine the parameters and outputs that will be exposed on the Shiny applications; and 

-   Select the appropriate Shiny UI components for exposing the parameters determined above. 

# Getting Started

## Loading R Packages

```{r}
pacman::p_load(tidyverse, lubridate, gridExtra, readxl, knitr, data.table, ggplot2, forecast, MLmetrics, tsbox, xts, plotly, hrbrthemes, imputeTS)
```

## Importing Data Into R

Talk about how we get this data: made use of data.gov.sg's API

```{r}
data <- read_csv("data/daily_historical.csv")
```

```{r}
glimpse(data)

```
# Temperature Data 

## Selecting the relevant columns for Temperature Data 
```{r}

temp <- data %>%
  select(station, year, month, day, mean_temperature, maximum_temperature, minimum_temperature) 

glimpse(temp)
```
## Creating a date column
```{r}
temp$tdate <- paste(temp$year, "-", temp$month, "-", temp$day)
glimpse(temp)
```

```{r}
temp <- temp %>%
  mutate(tdate = ymd(tdate))

glimpse(temp)
```
## Checking for missing values 

```{r}
summary(temp)

```

```{r}
temp <- temp %>%
  drop_na(c(year, month, day, tdate))

summary(temp)
```
```{r}
unique(temp$station)
```

## Further exploration of missing temperature using plotly


```{r}
weatherstations <- c("Macritchie Reservoir", "Lower Peirce Reservoir",  "Admiralty","East Coast Parkway", "Ang Mo Kio","Newton", "Lim Chu Kang","Marine Parade", "Choa Chu Kang (Central)", "Tuas South" , "Pasir Panjang", "Jurong Island", "Nicoll Highway","Botanic Garden", "Choa Chu Kang (South)","Whampoa", "Changi", "Jurong Pier",  "Ulu Pandan", "Mandai", "Tai Seng", "Jurong (West)", "Clementi", "Sentosa Island", "Bukit Panjang","Kranji Reservoir", "Upper Peirce Reservoir", "Kent Ridge", "Queenstown", "Tanjong Katong", "Somerset (Road)", "Punggol", "Simei", "Toa Payoh", "Tuas", "Bukit Timah","Pasir Ris (Central)")

temp_mean_wide <- temp %>%
  filter(year > 1983 & station %in% weatherstations) %>%
  select(tdate, station, mean_temperature) %>%
  pivot_wider(names_from = station, values_from = mean_temperature)

glimpse(temp_mean_wide)
```


```{r}
plot_ly(data = temp_mean_wide, 
        x = ~tdate, 
        y = ~ Admiralty, 
        type = "scatter", 
        mode = "lines+markers") |> 
  layout(title = "Temperature observed by Weather Station", 
       xaxis = list(title = "Date"), 
       yaxis = list(title = "", range = c(0,40)), 
      theme_ipsum_rc(plot_title_size = 13, plot_title_margin=4, subtitle_size=11, subtitle_margin=4,  
                 axis_title_size = 8, axis_text_size=8, axis_title_face= "bold", plot_margin = margin(4, 4, 4, 4)),  
       updatemenus = list(list(type = 'dropdown', 
                               xref = "paper", 
                               yref = "paper", 
                               xanchor = "left",
                               x = 0.04,
                               y = 0.95, 
                               buttons = list(
                                 list(method = "update",
                                      args = list(list(y = list(temp_mean_wide$Admiralty)), 
                                                    list(yaxis = list(title = "Temperature in Admiralty", range = c(0,40)))),label = "Admiralty"),
                                 list(method = "update",
                                      args = list(list(y = list(temp_mean_wide$`East Coast Parkway`)), 
                                                    list(yaxis = list(title = "Temperature in East Coast Parkway", range = c(0,40)))),label = "East Coast Parkway"), 
                                 list(method = "update",
                                      args = list(list(y = list(temp_mean_wide$`Ang Mo Kio`)), 
                                                    list(yaxis = list(title = "Temperature in Ang Mo Kio", range = c(0,40)))),label = "Ang Mo Kio"), 
                                 list(method = "update",
                                      args = list(list(y = list(temp_mean_wide$Newton)), 
                                                    list(yaxis = list(title = "Temperature in Newton", range = c(0,40)))),label = "Newton"), 
                                 list(method = "update",
                                      args = list(list(y = list(temp_mean_wide$`Tuas South`)), 
                                                    list(yaxis = list(title = "Temperature in Tuas South", range = c(0,40)))),label = "Tuas South"),
                                  list(method = "update",
                                      args = list(list(y = list(temp_mean_wide$`Pasir Panjang`)), 
                                                    list(yaxis = list(title = "Temperature in Pasir Panjang", range = c(0,40)))),label = "Pasir Panjang"), 
                                  list(method = "update",
                                      args = list(list(y = list(temp_mean_wide$`Jurong Island`)), 
                                                    list(yaxis = list(title = "Temperature in Jurong Island", range = c(0,40)))),label = "Jurong Island"), 
                                 list(method = "update",
                                      args = list(list(y = list(temp_mean_wide$`Choa Chu Kang (South)`)), 
                                                    list(yaxis = list(title = "Temperature in Choa Chu Kang (South)", range = c(0,40)))),label = "Choa Chu Kang (South)"), 
                                 list(method = "update", 
                                        args = list(list(y = list(temp_mean_wide$Changi)), 
                                                    list(yaxis = list(title = "Temperature in Changi", range = c(0,40)))),label = "Changi"),
                                  list(method = "update",
                                      args = list(list(y = list(temp_mean_wide$`Tai Seng`)), 
                                                    list(yaxis = list(title = "Temperature in Tai Seng", range = c(0,40)))),label = "Tai Seng"),
                                  list(method = "update",
                                      args = list(list(y = list(temp_mean_wide$`Jurong (West)`)), 
                                                    list(yaxis = list(title = "Temperature in Jurong West", range = c(0,40)))),label = "Jurong West"), 
                                   list(method = "update", 
                                        args = list(list(y = list(temp_mean_wide$Clementi)), 
                                                    list(yaxis = list(title = "Temperature  in Clementi", range = c(0,40)))),label = "Clementi"), 
                                   list(method = "update", 
                                        args = list(list(y = list(temp_mean_wide$`Sentosa Island`)), 
                                                    list(yaxis = list(title = "Temperature  in Sentosa", range = c(0,40)))),label = "Sentosa"), 
                                 list(method = "update", 
                                        args = list(list(y = list(temp_mean_wide$`Macritchie Reservoir`)), 
                                                    list(yaxis = list(title = "Temperature  at Macritchie Reservoir", range = c(0,40)))),label = "Macritchie Reservoir"), 
                                list(method = "update", 
                                        args = list(list(y = list(temp_mean_wide$`Lower Peirce Reservoir`)), 
                                                    list(yaxis = list(title = "Temperature  at Lower Peirce Reservoir", range = c(0,40)))),label = "Lower Peirce Reservoir"),
                                 list(method = "update", 
                                        args = list(list(y = list(temp_mean_wide$`Lim Chu Kang`)), 
                                                    list(yaxis = list(title = "Temperature at Lim Chu Kang", range = c(0,40)))),label = "Lim Chu Kang"),
                                list(method = "update", 
                                        args = list(list(y = list(temp_mean_wide$`Marine Parade`)), 
                                                    list(yaxis = list(title = "Temperature at Marine Parade", range = c(0,40)))),label = "Marine Parade"),
                                list(method = "update", 
                                        args = list(list(y = list(temp_mean_wide$`Choa Chu Kang (Central)`)), 
                                                    list(yaxis = list(title = "Temperature at Choa Chu Kang (Central)", range = c(0,40)))),label = "Choa Chu Kang (Central)"),
                                list(method = "update", 
                                        args = list(list(y = list(temp_mean_wide$`Choa Chu Kang (Central)`)), 
                                                    list(yaxis = list(title = "Temperature at Choa Chu Kang (Central)", range = c(0,40)))),label = "Choa Chu Kang (Central)"),
                                list(method = "update", 
                                        args = list(list(y = list(temp_mean_wide$`Nicoll Highway`)), 
                                                    list(yaxis = list(title = "Temperature at Nicoll Highway", range = c(0,40)))),label = "Nicoll Highway"),
                                list(method = "update", 
                                        args = list(list(y = list(temp_mean_wide$`Botanic Garden`)), 
                                                    list(yaxis = list(title = "Temperature at Botanic Garden", range = c(0,40)))),label = "Botanic Garden"),
                                list(method = "update", 
                                        args = list(list(y = list(temp_mean_wide$Whampoa)), 
                                                    list(yaxis = list(title = "Temperature at Whampoa", range = c(0,40)))),label = "Whampoa"),
                                list(method = "update", 
                                        args = list(list(y = list(temp_mean_wide$`Jurong Pier`)), 
                                                    list(yaxis = list(title = "Temperature at Jurong Pier", range = c(0,40)))),label = "Jurong Pier"),
                                list(method = "update", 
                                        args = list(list(y = list(temp_mean_wide$`Ulu Pandan`)), 
                                                    list(yaxis = list(title = "Temperature at Ulu Pandan", range = c(0,40)))),label = "Ulu Pandan"),
                                list(method = "update", 
                                        args = list(list(y = list(temp_mean_wide$Mandai)), 
                                                    list(yaxis = list(title = "Temperature at Mandai", range = c(0,40)))),label = "Mandai"),
                                list(method = "update", 
                                        args = list(list(y = list(temp_mean_wide$`Bukit Panjang`)), 
                                                    list(yaxis = list(title = "Temperature at Bukit Panjang", range = c(0,40)))),label = "Bukit Panjang"),
                                list(method = "update", 
                                        args = list(list(y = list(temp_mean_wide$`Kranji Reservoir`)), 
                                                    list(yaxis = list(title = "Temperature at Kranji Reservoir", range = c(0,40)))),label = "Kranji Reservoir"),
                                list(method = "update", 
                                        args = list(list(y = list(temp_mean_wide$`Upper Peirce Reservoir`)), 
                                                    list(yaxis = list(title = "Temperature at Upper Peirce Reservoir", range = c(0,40)))),label = "Upper Peirce Reservoir"),
                                list(method = "update", 
                                        args = list(list(y = list(temp_mean_wide$`Kent Ridge`)), 
                                                    list(yaxis = list(title = "Temperature at Kent Ridge", range = c(0,40)))),label = "Kent Ridge"),
                                list(method = "update", 
                                        args = list(list(y = list(temp_mean_wide$Queenstown)), 
                                                    list(yaxis = list(title = "Temperature at Queenstown", range = c(0,40)))),label = "Queenstown"),
                                list(method = "update", 
                                        args = list(list(y = list(temp_mean_wide$`Tanjong Katong`)), 
                                                    list(yaxis = list(title = "Temperature at Tanjong Katong", range = c(0,40)))),label = "Tanjong Katong"),
                                list(method = "update", 
                                        args = list(list(y = list(temp_mean_wide$`Somerset (Road)`)), 
                                                    list(yaxis = list(title = "Temperature at Somerset (Road)", range = c(0,40)))),label = "Somerset (Road)"),
                                list(method = "update", 
                                        args = list(list(y = list(temp_mean_wide$`Punggol`)), 
                                                    list(yaxis = list(title = "Temperature at Punggol", range = c(0,40)))),label = "Punggol"),
                                list(method = "update", 
                                        args = list(list(y = list(temp_mean_wide$`Simen`)), 
                                                    list(yaxis = list(title = "Temperature at Simei", range = c(0,40)))),label = "Simei"),
                                list(method = "update", 
                                        args = list(list(y = list(temp_mean_wide$`Toa Payoh`)), 
                                                    list(yaxis = list(title = "Temperature at Toa Payoh", range = c(0,40)))),label = "Toa Payoh"),
                                list(method = "update", 
                                        args = list(list(y = list(temp_mean_wide$`Tuas`)), 
                                                    list(yaxis = list(title = "Temperature at Tuas", range = c(0,40)))),label = "Tuas"),
                                list(method = "update", 
                                        args = list(list(y = list(temp_mean_wide$`Bukit Timah`)), 
                                                    list(yaxis = list(title = "Temperature at Bukit Timah", range = c(0,40)))),label = "Bukit Timah"),
                                list(method = "update", 
                                        args = list(list(y = list(temp_mean_wide$`Pasir Ris (Central)`)), 
                                                    list(yaxis = list(title = "Temperature at Pasir Ris (Central)", range = c(0,40)))),label = "Pasir Ris (Central)")
                               ))))  



```


Seems like there are some weather stations with no temperature data, some weather stations have temperature data for certain years, and some stations have temperature data from 1984 onwards. 

Let's explore further.

```{r}
missing.values <- temp_mean_wide %>%
  gather(key = "key", value = "val") %>%
  mutate(isna = is.na(val)) %>%
  group_by(key) %>%
  mutate(total = n()) %>%
  group_by(key, total, isna) %>%
  summarise(num.isna = n()) %>%
  mutate(pct = num.isna / total * 100)

levels <-
    (missing.values  %>% filter(isna == T) %>% arrange(desc(pct)))$key

percentage.plot <- missing.values %>%
      ggplot() +
        geom_bar(aes(x = reorder(key, desc(pct)), 
                     y = pct, fill=isna), 
                 stat = 'identity', alpha=0.8) +
      scale_x_discrete(limits = levels) +
      scale_fill_manual(name = "", 
                        values = c('steelblue', 'tomato3'), labels = c("Present", "Missing")) +
      coord_flip() +
      labs(title = "Percentage of missing values", x =
             'Variable', y = "% of missing values")

percentage.plot


```

Drop the cols with missing values 

```{r}
temp_nomissing <- temp %>%
  drop_na()

glimpse(temp_nomissing)
```

```{r}
unique(temp_nomissing$station)
```

```{r}

temp_mean_widec <- temp_nomissing %>%
  filter(year > 1983) %>%
  select(tdate, station, mean_temperature) %>%
  pivot_wider(names_from = station, values_from = mean_temperature)

glimpse(temp_mean_wide)
```
```{r}
plot_ly(data = temp_mean_widec, 
        x = ~tdate, 
        y = ~ Admiralty, 
        type = "scatter", 
        mode = "lines+markers") |> 
  layout(title = "Temperature observed by Weather Station", 
       xaxis = list(title = "Date"), 
       yaxis = list(title = "", range = c(0,40)), 
      theme_ipsum_rc(plot_title_size = 13, plot_title_margin=4, subtitle_size=11, subtitle_margin=4,  
                 axis_title_size = 8, axis_text_size=8, axis_title_face= "bold", plot_margin = margin(4, 4, 4, 4)),  
       updatemenus = list(list(type = 'dropdown', 
                               xref = "paper", 
                               yref = "paper", 
                               xanchor = "left",
                               x = 0.04,
                               y = 0.95, 
                               buttons = list(
                                 list(method = "update",
                                      args = list(list(y = list(temp_mean_widec$Admiralty)), 
                                                    list(yaxis = list(title = "Temperature in Admiralty", range = c(0,40)))),label = "Admiralty"),
                                 list(method = "update",
                                      args = list(list(y = list(temp_mean_widec$`East Coast Parkway`)), 
                                                    list(yaxis = list(title = "Temperature in East Coast Parkway", range = c(0,40)))),label = "East Coast Parkway"), 
                                 list(method = "update",
                                      args = list(list(y = list(temp_mean_widec$`Ang Mo Kio`)), 
                                                    list(yaxis = list(title = "Temperature in Ang Mo Kio", range = c(0,40)))),label = "Ang Mo Kio"), 
                                 list(method = "update",
                                      args = list(list(y = list(temp_mean_widec$Newton)), 
                                                    list(yaxis = list(title = "Temperature in Newton", range = c(0,40)))),label = "Newton"), 
                                 list(method = "update",
                                      args = list(list(y = list(temp_mean_widec$`Tuas South`)), 
                                                    list(yaxis = list(title = "Temperature in Tuas South", range = c(0,40)))),label = "Tuas South"),
                                  list(method = "update",
                                      args = list(list(y = list(temp_mean_widec$`Pasir Panjang`)), 
                                                    list(yaxis = list(title = "Temperature in Pasir Panjang", range = c(0,40)))),label = "Pasir Panjang"), 
                                  list(method = "update",
                                      args = list(list(y = list(temp_mean_widec$`Jurong Island`)), 
                                                    list(yaxis = list(title = "Temperature in Jurong Island", range = c(0,40)))),label = "Jurong Island"), 
                                 list(method = "update",
                                      args = list(list(y = list(temp_mean_widec$`Choa Chu Kang (South)`)), 
                                                    list(yaxis = list(title = "Temperature in Choa Chu Kang", range = c(0,40)))),label = "Choa Chu Kang"), 
                                 list(method = "update", 
                                        args = list(list(y = list(temp_mean_widec$Changi)), 
                                                    list(yaxis = list(title = "Temperature in Changi", range = c(0,40)))),label = "Changi"),
                                  list(method = "update",
                                      args = list(list(y = list(temp_mean_widec$`Tai Seng`)), 
                                                    list(yaxis = list(title = "Temperature in Tai Seng", range = c(0,40)))),label = "Tai Seng"),
                                  list(method = "update",
                                      args = list(list(y = list(temp_mean_widec$`Jurong (West)`)), 
                                                    list(yaxis = list(title = "Temperature in Jurong West", range = c(0,40)))),label = "Jurong West"), 
                                   list(method = "update", 
                                        args = list(list(y = list(temp_mean_widec$Clementi)), 
                                                    list(yaxis = list(title = "Temperature  in Clementi", range = c(0,40)))),label = "Clementi"), 
                                   list(method = "update", 
                                        args = list(list(y = list(temp_mean_widec$`Sentosa Island`)), 
                                                    list(yaxis = list(title = "Temperature  in Sentosa", range = c(0,40)))),label = "Sentosa")
                                   
                               ))))  



```

There are some missing time gaps in the data. 


## Preparing for time series forecasting 



```{r}
changi <- temp %>%
  filter(station == "Changi")

glimpse(changi)
```

```{r}
adm <- temp %>%
  filter(station == "Admiralty")

glimpse(adm)
```


```{r}
adm_meantemp <- adm %>% 
  select(date, mean_temperature)

ggplot_na_distribution(adm_meantemp)
```


```{r}
changi <- changi %>%
  drop_na() %>%
  select(date, mean_temperature, maximum_temperature, minimum_temperature)

glimpse(changi)
```

```{r}
range(changi$date)
```


```{r}
changits <- xts(changi[,c("mean_temperature","maximum_temperature","minimum_temperature")], order.by=as.Date(changi$date))

```

```{r}
ggplot_na_distribution2(changits$minimum_temperature)
```

```{r}
changits <- ts_regular(changits)
changits <- na.fill(changits, "extend")
changits <- window(changits, start = as.Date("1982-01-01"), end = as.Date("2023-12-31"))

```

```{r}
plot(ts_ts(changits$maximum_temperature), col="darkred", bty="n", las=1, fg=NA, 
	ylim=c(20, 40), ylab="Temperature (C)")

lines(ts_ts(changits$minimum_temperature), col="navy")

lines(ts_ts(changits$mean_temperature), col="darkgreen")

grid(nx=NA, ny=NULL, lty=1, col="gray")

legend("topright", fill=c("darkred", "darkgreen", "navy"), cex=0.7,
	legend=c("MAX", "MEAN", "MIN"), bg="white")


```

```{r}

summary(changits)

```

```{r}

changits$MONTH <- format(index(changits), "%m")

months <- split(as.numeric(changits$maximum_temperature), changits$MONTH)

sapply(months, summary)
```


```{r}
decomposition = stl(ts_ts(changits$maximum_temperature), s.window=365, t.window=7001)

plot(decomposition)

summary(decomposition$time.series[,"trend"])


```


```{r}
monthly.tmax <- period.apply(changits$maximum_temperature, INDEX = seq(1, nrow(changits) - 1, 30.4375), FUN = colMeans)

plot(ts_ts(monthly.tmax), col="darkred", ylim=c(20, 40), 
	lwd=3, bty="n", las=1, fg=NA, ylab="TMAX (C)")

grid(nx=NA, ny=NULL, lty=1)


```
# Time Series Forecasting for Max Temp    

## Holt-Winters 

```{r}

training.data = period.apply(changits$maximum_temperature, seq(1, nrow(changits) - 1, 30.4375), max)

model.tmax = hw(ts_ts(training.data), h=60)

```

```{r}

plot(model.tmax, lwd=3, bty="n", las=1, fg=NA)

grid(nx=NA, ny=NULL, lty=1)

```
## ARIMA 

```{r}
training.data <- ts_ts(changits$maximum_temperature)

parameters <- auto.arima(training.data)

print(parameters)

```

```{r}

#arima.model = arima(training.data, order = c(4,1,1), seasonal = list(order = c(4,1,1), period = 365))

arima.tmax = forecast(parameters, h = 10)

plot(arima.tmax, lwd=3, bty="n", las=1, fg=NA, 
	xlim=c(1982, 2030), ylab="Mean Monthly High Temperature (C)")

grid(nx=NA, ny=NULL, lty=1)

```

# Data Exploration using plotly 

```{r}
tempstations <- c("Admiralty", "Ang Mo Kio", "Changi", "Choa Chu Kang (South)", "Clementi", "East Coast Parkway",
                      "Jurong (West)", "Jurong Island", "Newton", "Pasir Panjang", "Paya Lebar", 
                      "Pulau Ubin", "Sembawang", "Sentosa Island", "Tai Seng", "Tuas South")
temp_mean_wide <- temp %>%
  filter(year > 1983 & station %in% tempstations) %>%
  select(date, station, mean_temperature) %>%
  pivot_wider(names_from = station, values_from = mean_temperature)

glimpse(temp_mean_wide)
```


```{r}
ggplot_na_distribution(temp_mean_wide)


```

```{r}
plot_ly(data = temp_mean_wide, 
        x = ~date, 
        y = ~ Admiralty, 
        type = "scatter", 
        mode = "lines+markers") |> 
  layout(title = "Temperature observed by Weather Station", 
       xaxis = list(title = "Date"), 
       yaxis = list(title = "", range = c(0,40)), 
      theme_ipsum_rc(plot_title_size = 13, plot_title_margin=4, subtitle_size=11, subtitle_margin=4,  
                 axis_title_size = 8, axis_text_size=8, axis_title_face= "bold", plot_margin = margin(4, 4, 4, 4)),  
       updatemenus = list(list(type = 'dropdown', 
                               xref = "paper", 
                               yref = "paper", 
                               xanchor = "left",
                               x = 0.04,
                               y = 0.95, 
                               buttons = list(
                                 list(method = "update",
                                      args = list(list(y = list(temp_mean_wide$Admiralty)), 
                                                    list(yaxis = list(title = "Temperature in Admiralty", range = c(0,40)))),label = "Admiralty"),
                                 list(method = "update",
                                      args = list(list(y = list(temp_mean_wide$`East Coast Parkway`)), 
                                                    list(yaxis = list(title = "Temperature in East Coast Parkway", range = c(0,40)))),label = "East Coast Parkway"), 
                                 list(method = "update",
                                      args = list(list(y = list(temp_mean_wide$`Ang Mo Kio`)), 
                                                    list(yaxis = list(title = "Temperature in Ang Mo Kio", range = c(0,40)))),label = "Ang Mo Kio"), 
                                 list(method = "update",
                                      args = list(list(y = list(temp_mean_wide$Newton)), 
                                                    list(yaxis = list(title = "Temperature in Newton", range = c(0,40)))),label = "Newton"), 
                                 list(method = "update",
                                      args = list(list(y = list(temp_mean_wide$`Tuas South`)), 
                                                    list(yaxis = list(title = "Temperature in Tuas South", range = c(0,40)))),label = "Tuas South"),
                                  list(method = "update",
                                      args = list(list(y = list(temp_mean_wide$`Pasir Panjang`)), 
                                                    list(yaxis = list(title = "Temperature in Pasir Panjang", range = c(0,40)))),label = "Pasir Panjang"), 
                                  list(method = "update",
                                      args = list(list(y = list(temp_mean_wide$`Jurong Island`)), 
                                                    list(yaxis = list(title = "Temperature in Jurong Island", range = c(0,40)))),label = "Jurong Island"), 
                                 list(method = "update",
                                      args = list(list(y = list(temp_mean_wide$`Choa Chu Kang (South)`)), 
                                                    list(yaxis = list(title = "Temperature in Choa Chu Kang", range = c(0,40)))),label = "Choa Chu Kang"), 
                                 list(method = "update", 
                                        args = list(list(y = list(temp_mean_wide$Changi)), 
                                                    list(yaxis = list(title = "Temperature in Changi", range = c(0,40)))),label = "Changi"),
                                  list(method = "update",
                                      args = list(list(y = list(temp_mean_wide$`Tai Seng`)), 
                                                    list(yaxis = list(title = "Temperature in Tai Seng", range = c(0,40)))),label = "Tai Seng"),
                                  list(method = "update",
                                      args = list(list(y = list(temp_mean_wide$`Jurong (West)`)), 
                                                    list(yaxis = list(title = "Temperature in Jurong West", range = c(0,40)))),label = "Jurong West"), 
                                   list(method = "update", 
                                        args = list(list(y = list(temp_mean_wide$Clementi)), 
                                                    list(yaxis = list(title = "Temperature  in Clementi", range = c(0,40)))),label = "Clementi"), 
                                   list(method = "update", 
                                        args = list(list(y = list(temp_mean_wide$`Sentosa Island`)), 
                                                    list(yaxis = list(title = "Temperature  in Sentosa", range = c(0,40)))),label = "Sentosa")
                                   
                               ))))  



```




```{r}

temp_mean_widec <- temp_mean_wide %>%
  drop_na()

glimpse(temp_mean_widec)
```


```{r}
plot_ly(data = temp_mean_widec, 
        x = ~date, 
        y = ~ Admiralty, 
        type = "scatter", 
        mode = "lines+markers") |> 
  layout(title = "Temperature observed by Weather Station", 
       xaxis = list(title = "Date"), 
       yaxis = list(title = "", range = c(0,40)), 
      theme_ipsum_rc(plot_title_size = 13, plot_title_margin=4, subtitle_size=11, subtitle_margin=4,  
                 axis_title_size = 8, axis_text_size=8, axis_title_face= "bold", plot_margin = margin(4, 4, 4, 4)),  
       updatemenus = list(list(type = 'dropdown', 
                               xref = "paper", 
                               yref = "paper", 
                               xanchor = "left",
                               x = 0.04,
                               y = 0.95, 
                               buttons = list(
                                 list(method = "update",
                                      args = list(list(y = list(temp_mean_widec$Admiralty)), 
                                                    list(yaxis = list(title = "Temperature in Admiralty", range = c(0,40)))),label = "Admiralty"),
                                 list(method = "update",
                                      args = list(list(y = list(temp_mean_widec$`East Coast Parkway`)), 
                                                    list(yaxis = list(title = "Temperature in East Coast Parkway", range = c(0,40)))),label = "East Coast Parkway"), 
                                 list(method = "update",
                                      args = list(list(y = list(temp_mean_widec$`Ang Mo Kio`)), 
                                                    list(yaxis = list(title = "Temperature in Ang Mo Kio", range = c(0,40)))),label = "Ang Mo Kio"), 
                                 list(method = "update",
                                      args = list(list(y = list(temp_mean_widec$Newton)), 
                                                    list(yaxis = list(title = "Temperature in Newton", range = c(0,40)))),label = "Newton"), 
                                 list(method = "update",
                                      args = list(list(y = list(temp_mean_widec$`Tuas South`)), 
                                                    list(yaxis = list(title = "Temperature in Tuas South", range = c(0,40)))),label = "Tuas South"),
                                  list(method = "update",
                                      args = list(list(y = list(temp_mean_widec$`Pasir Panjang`)), 
                                                    list(yaxis = list(title = "Temperature in Pasir Panjang", range = c(0,40)))),label = "Pasir Panjang"), 
                                  list(method = "update",
                                      args = list(list(y = list(temp_mean_widec$`Jurong Island`)), 
                                                    list(yaxis = list(title = "Temperature in Jurong Island", range = c(0,40)))),label = "Jurong Island"), 
                                 list(method = "update",
                                      args = list(list(y = list(temp_mean_widec$`Choa Chu Kang (South)`)), 
                                                    list(yaxis = list(title = "Temperature in Choa Chu Kang", range = c(0,40)))),label = "Choa Chu Kang"), 
                                 list(method = "update", 
                                        args = list(list(y = list(temp_mean_widec$Changi)), 
                                                    list(yaxis = list(title = "Temperature in Changi", range = c(0,40)))),label = "Changi"),
                                  list(method = "update",
                                      args = list(list(y = list(temp_mean_widec$`Tai Seng`)), 
                                                    list(yaxis = list(title = "Temperature in Tai Seng", range = c(0,40)))),label = "Tai Seng"),
                                  list(method = "update",
                                      args = list(list(y = list(temp_mean_widec$`Jurong (West)`)), 
                                                    list(yaxis = list(title = "Temperature in Jurong West", range = c(0,40)))),label = "Jurong West"), 
                                   list(method = "update", 
                                        args = list(list(y = list(temp_mean_widec$Clementi)), 
                                                    list(yaxis = list(title = "Temperature  in Clementi", range = c(0,40)))),label = "Clementi"), 
                                   list(method = "update", 
                                        args = list(list(y = list(temp_mean_widec$`Sentosa Island`)), 
                                                    list(yaxis = list(title = "Temperature  in Sentosa", range = c(0,40)))),label = "Sentosa")
                                   
                               ))))  



```








# Rainfall Data 
save it into rds 

## Data Wrangling 

### Check for Missing Data 

### Check Data type 

## Time Series Forecasting 

# Wind Data 
save it into rds 

## Data Wrangling 


## Time Series Forecasting 





# Shiny Dashboard Prototype - UI and Server Design 

## UI 

### Sketch 

### Details 



## Server 

### Details 







# To adjust 

```{r}
missingv <- data %>% 
  map(is.na) %>%
  map(sum) 
  
missingv <- as_tibble(missingv)

kable(missingv)

```

```{r}

# Create a data frame with missing value counts
missing.values <- data %>%
    gather(key = "key", value = "val") %>%
    mutate(is.missing = is.na(val)) %>%
    group_by(key, is.missing) %>%
    summarise(num.missing = n()) %>%
    filter(is.missing==T) %>%
    select(-is.missing) %>%
    arrange(desc(num.missing)) 

# Create a bar chart
missing.values %>%
  ggplot() +
    geom_bar(aes(x=key, y=num.missing), stat = 'identity') +
    labs(x='variable', y="number of missing values", title='Number of missing values') +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

```{r}
missing.values <- data %>%
  gather(key = "key", value = "val") %>%
  mutate(isna = is.na(val)) %>%
  group_by(key) %>%
  mutate(total = n()) %>%
  group_by(key, total, isna) %>%
  summarise(num.isna = n()) %>%
  mutate(pct = num.isna / total * 100)

levels <-
    (missing.values  %>% filter(isna == T) %>% arrange(desc(pct)))$key

percentage.plot <- missing.values %>%
      ggplot() +
        geom_bar(aes(x = reorder(key, desc(pct)), 
                     y = pct, fill=isna), 
                 stat = 'identity', alpha=0.8) +
      scale_x_discrete(limits = levels) +
      scale_fill_manual(name = "", 
                        values = c('steelblue', 'tomato3'), labels = c("Present", "Missing")) +
      coord_flip() +
      labs(title = "Percentage of missing values", x =
             'Variable', y = "% of missing values")

percentage.plot


```

```{r}
row.plot <- data %>%
  mutate(id = row_number()) %>%
  gather(-id, key = "key", value = "val") %>%
  mutate(isna = is.na(val)) %>%
  ggplot(aes(key, id, fill = isna)) +
    geom_raster(alpha=0.8) +
    scale_fill_manual(name = "",
        values = c('steelblue', 'tomato3'),
        labels = c("Present", "Missing")) +
    scale_x_discrete(limits = levels) +
    labs(x = "Variable",
           y = "Row Number", title = "Missing values in rows") +
    coord_flip()

row.plot
```










```{r}
data_nomissing <- drop_na(data)

```

Confirm that there is no missing values

```{r}

# Create a data frame with missing value counts
missing.values2 <- data_nomissing %>%
    gather(key = "key", value = "val") %>%
    mutate(is.missing = is.na(val)) %>%
    group_by(key, is.missing) %>%
    summarise(num.missing = n()) %>%
    filter(is.missing==T) %>%
    select(-is.missing) %>%
    arrange(desc(num.missing)) 

# Create a bar chart
missing.values2 %>%
  ggplot() +
    geom_bar(aes(x=key, y=num.missing), stat = 'identity') +
    labs(x='variable', y="number of missing values", title='Number of missing values') +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

select temperature data

```{r}

meantemp <- data_nomissing %>%
  select(station, year, month, day, mean_temperature) 

glimpse(meantemp)
```

```{r}

meantemp$date <- paste(meantemp$year, "-", meantemp$month, "-", meantemp$day)
glimpse(meantemp)
```

```{r}
meantemp <- meantemp %>%
  mutate(date = ymd(date))

glimpse(meantemp)
```

```{r}

unique(meantemp$station)
```
